---
title: "1. Tidy data: trial- and child-level frames"
subtitle: "peekbank-items"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

Builds the trial-level (`1_d_trial`) and child-level (`1_d_sub`) data frames
from the cached AOI data. Adapted from `peekbank-development/1_tidy_data.qmd`.

```{r setup}
library(here)
source(here("helper", "common.R"))
source(here("helper", "rt_helper.R"))

ACC_MIN_TRIAL_CUTOFF <- 4
RT_MIN_TRIAL_CUTOFF <- 2

d_aoi <- readRDS(here("cached_intermediates", "0_d_aoi.Rds"))
```

## Trial-level accuracy

Two windows: classic short (200–2000 ms) and ICC-maximizing long (200–4000 ms).

```{r trial-acc}
d_trial <- d_aoi |>
  mutate(short_window = t_norm > 200 & t_norm <= 2000,
         long_window = t_norm > 200 & t_norm <= 4000) |>
  group_by(dataset_name, dataset_id, administration_id, subject_id,
           trial_id, age, target_label, distractor_label) |>
  summarise(
    short_window_accuracy = mean(correct[short_window], na.rm = TRUE),
    long_window_accuracy = mean(correct[long_window], na.rm = TRUE),
    short_window_prop_data = sum(!is.na(correct[short_window]), na.rm = TRUE) /
      length(correct[short_window]),
    long_window_prop_data = sum(!is.na(correct[long_window]), na.rm = TRUE) /
      length(correct[long_window]),
    short_window_target = sum(correct[short_window], na.rm = TRUE),
    long_window_target = sum(correct[long_window], na.rm = TRUE),
    short_window_distractor = sum(!correct[short_window], na.rm = TRUE),
    long_window_distractor = sum(!correct[long_window], na.rm = TRUE)
  ) |>
  mutate(
    short_window_elogit = log((short_window_target + .5) / (short_window_distractor + .5)),
    long_window_elogit = log((long_window_target + .5) / (long_window_distractor + .5))
  )
```

## Reaction times

Computed from run-length-encoded AOI sequences (the shiny-app workflow).

```{r rt}
d_rle <- d_aoi |>
  filter(any(t_norm == 0), t_norm >= 0) |>
  group_by(administration_id, trial_id, trial_order) |>
  reframe(lengths = rle(aoi)$lengths, values = rle(aoi)$values)

d_rt <- d_rle |>
  group_by(administration_id, trial_id, trial_order) |>
  nest() |>
  mutate(data = lapply(data, get_rt)) |>
  unnest(cols = c(data))

d_trial_rt_unfiltered <- d_trial |>
  left_join(d_rt) |>
  mutate(
    raw_rt = rt,
    rt = case_when(
      dataset_name == "borovsky_2019" ~ NA,  # gaze-contingent; RTs not comparable
      shift_type == "D-T" ~ rt,
      TRUE ~ NA
    ),
    log_rt = log(rt)
  )
```

## Filtering

Remove trials with <50% data in both windows; NA-out window measures with <50%
data; drop RTs < 367 ms.

```{r filter}
d_trial <- d_trial_rt_unfiltered |>
  filter(!(is.na(short_window_prop_data) & is.na(long_window_prop_data)),
         !(short_window_prop_data < .5 & long_window_prop_data < .5)) |>
  mutate(across(contains("window"), ~ ifelse(is.na(.x), NA, .x))) |>
  mutate(
    short_window_accuracy = ifelse(short_window_prop_data >= .5, short_window_accuracy, NA),
    long_window_accuracy = ifelse(long_window_prop_data >= .5, long_window_accuracy, NA),
    rt = ifelse(short_window_prop_data >= .5, rt, NA),
    short_window_elogit = ifelse(short_window_prop_data >= .5, short_window_elogit, NA),
    long_window_elogit = ifelse(long_window_prop_data >= .5, long_window_elogit, NA),
    rt = ifelse(rt < 367, NA, rt),
    log_rt = ifelse(rt < 367, NA, log_rt)
  )
```

## Descriptives

```{r desc-overall}
d_trial |>
  ungroup() |>
  summarise(
    n = n(),
    avg_acc_short = mean(short_window_accuracy, na.rm = TRUE),
    avg_acc_long = mean(long_window_accuracy, na.rm = TRUE),
    avg_rt = mean(rt, na.rm = TRUE),
    nas_in_acc_short = mean(is.na(short_window_accuracy)),
    nas_in_acc_long = mean(is.na(long_window_accuracy)),
    prop_rts = mean(!is.na(rt))
  ) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  kable(digits = 2)
```

```{r desc-dataset}
d_trial |>
  group_by(dataset_name) |>
  summarise(
    n = n(),
    avg_acc_short = mean(short_window_accuracy, na.rm = TRUE),
    avg_acc_long = mean(long_window_accuracy, na.rm = TRUE),
    avg_rt = mean(rt, na.rm = TRUE),
    prop_rts = mean(!is.na(rt))
  ) |>
  arrange(desc(n)) |>
  kable(digits = 2)
```

```{r hist-acc}
ggplot(d_trial, aes(x = long_window_accuracy)) +
  geom_histogram(bins = 30) +
  facet_wrap(~dataset_name, scales = "free_y") +
  xlab("Long-window accuracy")
```

## Child-level frame

```{r sub}
d_sub_nocdi <- d_trial |>
  group_by(dataset_name, subject_id, administration_id, age) |>
  summarise(
    n_trials = sum(!is.na(long_window_accuracy)),
    n_trials_rt = sum(!is.na(rt)),
    rt_var = sd(rt, na.rm = TRUE),
    log_rt_var = sd(log_rt, na.rm = TRUE),
    short_window_acc_var = sd(short_window_accuracy, na.rm = TRUE),
    long_window_acc_var = sd(long_window_accuracy, na.rm = TRUE),
    across(c(contains("_window"), "rt", "log_rt"), ~ mean(.x, na.rm = TRUE))
  )

# apply trial-count exclusions
d_sub_nocdi_exclusions <- d_sub_nocdi |>
  mutate(
    across(c("rt", "rt_var"), ~ ifelse(n_trials_rt < RT_MIN_TRIAL_CUTOFF, NA, .x)),
    across(contains("log_rt"), ~ ifelse(n_trials_rt < RT_MIN_TRIAL_CUTOFF, NA, .x)),
    across(contains("window"), ~ ifelse(n_trials < ACC_MIN_TRIAL_CUTOFF, NA, .x))
  )

# join CDI summary scores
cdi_data <- readRDS(here("cached_intermediates", "0_cdi_subjects.Rds"))
d_sub <- d_sub_nocdi_exclusions |>
  left_join(select(cdi_data, -age, -dataset_name),
            by = c("subject_id", "administration_id")) |>
  mutate(log_age = log(age)) |>
  ungroup()
```

```{r sub-desc}
print(paste(length(unique(d_sub$subject_id)), "children"))
print(paste(length(unique(d_sub$administration_id)), "administrations"))
print(paste(sum(!is.na(d_sub$prod) | !is.na(d_sub$comp)), "administrations have CDI data"))
print(paste(length(unique(d_sub$dataset_name)), "datasets"))
```

## Save

```{r save}
saveRDS(d_trial, here("cached_intermediates", "1_d_trial.Rds"))
write_csv(d_trial, here("cached_intermediates", "1_d_trial.csv"), na = "")
saveRDS(d_sub, here("cached_intermediates", "1_d_sub.Rds"))
write_csv(d_sub, here("cached_intermediates", "1_d_sub.csv"), na = "")
```
