---
title: "4. Aggregate item-level linking (the classic analysis, updated)"
subtitle: "peekbank-items"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

This notebook reproduces the **background failure** (`4A`: naive word-level LWL
accuracy correlates with Wordbank AoA in the *wrong* direction) on 2026.1 data,
then tests whether the two methodological fixes help **at the aggregate item
level**:

1. a better LWL measure (animacy-adjusted random intercept vs raw accuracy);
2. a better CDI measure (IRT difficulty `b` vs AoA).

No child information enters here — this is the "aggregate" data setting. The
trial-level child×item comparison across settings is in notebook 5.

```{r setup}
library(here)
library(ggrepel)
source(here("helper", "common.R"))
knitr::opts_chunk$set(cache = FALSE)

lwl <- readRDS(here("cached_intermediates", "2_lwl_item.Rds"))
cdi <- readRDS(here("cached_intermediates", "3_cdi_difficulty.Rds"))

# CDI-side difficulty per word
cdi_diff <- cdi$item_pars |>
  select(item, a, b) |>
  left_join(cdi$aoas |> select(item, aoa), by = "item")

# LWL-side measures per word, merged with CDI difficulty
mdf <- lwl$lwl_word |>
  mutate(item = target_label) |>
  inner_join(cdi_diff, by = "item")
```

We have `r nrow(mdf)` words with both LWL and CDI measures.

## The naive failure: raw accuracy vs AoA

Direction we *want*: earlier-acquired (low AoA) words should be recognized
**better** (high accuracy) → a **negative** slope.

```{r naive}
plot_corr <- function(data, x, y, xlab, ylab) {
  r <- cor(data[[x]], data[[y]], use = "complete.obs")
  ggplot(data, aes(x = .data[[x]], y = .data[[y]])) +
    geom_point(aes(size = n_trials), alpha = .4) +
    geom_smooth(method = "lm") +
    geom_text_repel(aes(label = target_label), size = 2.3, max.overlaps = 8) +
    xlab(xlab) + ylab(ylab) +
    labs(subtitle = sprintf("r = %.2f", r)) +
    theme(legend.position = "none")
}

plot_corr(mdf, "aoa", "avg_accuracy", "Wordbank AoA (months)", "Avg accuracy")
```

As in the background analysis, this goes the **wrong way**: later-acquired words
show *higher* accuracy, driven by age confounding (easy words are tested in
younger samples) and salience.

## Fix 1: animacy-adjusted LWL difficulty

```{r fix-lwl}
plot_corr(mdf, "aoa", "intercept_adj",
          "Wordbank AoA (months)", "LWL ease (animacy-adjusted intercept)")
```

## Fix 2: IRT difficulty instead of AoA

```{r fix-cdi}
plot_corr(mdf, "b", "intercept_adj",
          "CDI IRT difficulty (b)", "LWL ease (animacy-adjusted intercept)")
```

## Summary of correlations

All cells are correlations of an LWL measure with a CDI difficulty measure.
**Success = negative** (harder/later word ⇒ lower LWL ease). The naive cell
(top-left, `avg_accuracy` × `aoa`) is the one the background found positive.

```{r summary-table}
lwl_measures <- c("avg_accuracy", "avg_elogit", "intercept_base", "intercept_adj")
cdi_measures <- c("aoa", "b")

grid <- expand_grid(lwl = lwl_measures, cdi = cdi_measures, hf_only = c(FALSE, TRUE))

results <- grid |>
  rowwise() |>
  mutate(
    r = {
      d <- if (hf_only) filter(mdf, n_trials > 200) else mdf
      cor(d[[lwl]], d[[cdi]], use = "complete.obs")
    },
    n_words = if (hf_only) sum(mdf$n_trials > 200) else nrow(mdf)
  ) |>
  ungroup()

results |>
  mutate(cell = paste0(cdi, ifelse(hf_only, " (n>200)", ""))) |>
  select(lwl, cell, r) |>
  pivot_wider(names_from = cell, values_from = r) |>
  kable(digits = 2, caption = "Correlation of LWL measure (rows) with CDI difficulty (cols). Negative = success.")
```

## Read

The table makes the diagnosis concrete: which combination of LWL measure and CDI
measure (and frequency filter) is needed before the aggregate item-level
correlation even points the right way. Whatever the verdict, the aggregate level
throws away the within-child structure — a child only "knows" a word
probabilistically, and the 2AFC decision depends on the *distractor* too. That
structure is the subject of notebook 5.
