---
title: "5. Linking hypotheses × data settings"
subtitle: "peekbank-items"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

We want to know whether a child's **word knowledge** predicts their **trial-level
looking accuracy**, and which form of that link is best supported. This notebook
sets up the comparison cleanly: a few explicit **linking models**, each fit under
each available **way of measuring word knowledge** (a "data setting"), reported
three ways — as coefficients, as accuracy effect sizes, and as out-of-sample
prediction.

Notebook 6 then sharpens the winning idea into the strict item-specific test.

## The dependent variable and what "tiny R²" means

The outcome is the trial **elogit**, `log((target looks + .5)/(distractor looks
+ .5))`. Conveniently, `plogis(elogit)` ≈ the trial's looking accuracy, so we can
translate any elogit effect into **percentage points of accuracy**.

A single trial is a *very* noisy measurement — a few seconds of a toddler's
gaze. So most trial-level variance is irreducible noise, and **no** model can
explain much of it. We make that ceiling explicit below; it is why the
out-of-sample R² values are all near 0.03–0.04. The right yardsticks are
therefore (a) the **coefficient / accuracy effect size** and (b) the
**improvement over an age-only baseline**, not the absolute R².

```{r setup}
library(here)
library(lme4)
source(here("helper", "common.R"))
knitr::opts_chunk$set(cache = FALSE)

d_trial <- readRDS(here("cached_intermediates", "1_d_trial.Rds")) |> ungroup()
d_sub   <- readRDS(here("cached_intermediates", "1_d_sub.Rds"))
cdi     <- readRDS(here("cached_intermediates", "3_cdi_difficulty.Rds"))
subset_dat  <- readRDS(here("cached_intermediates", "0_cdi_items_subset.Rds"))
subject_map <- readRDS(here("cached_intermediates", "0_subject_map.Rds"))

p2pl <- function(theta, a, b) plogis(a * (theta - b))
```

## Three ways to measure word knowledge ("data settings")

For each trial we form a **target-knowledge** signal `k_t` and a
**distractor-knowledge** signal `k_d`, oriented so that more knowledge = larger
value. The three settings differ in where that knowledge comes from:

| setting | `k_t`, `k_d` | child-specific? | item-specific? | covers |
|---|---|---|---|---|
| **aggregate** | `-b` (IRT *ease* of the word) | no | yes | all trials |
| **imputed** | `P(knows) = 2PL(θ_child, a, b)` | yes (via CDI total) | only via difficulty | CDI children |
| **precise** | child *actually produced* the word (0/1) | yes | yes (real) | TotLot subset |

The imputed signal is built from the child's overall CDI ability `θ` and the
word's IRT parameters; the precise signal is the child's real per-word CDI
response.

```{r build-frame}
cdi_diff <- cdi$item_pars |> select(item, a, b)

base <- d_trial |>
  rename(accuracy = long_window_accuracy, elogit = long_window_elogit) |>
  filter(!is.na(elogit)) |>
  mutate(log_age = log(age), log_age_centered = log_age - mean(log_age)) |>
  left_join(cdi_diff |> rename(a_t = a, b_t = b), by = c("target_label" = "item")) |>
  left_join(cdi_diff |> rename(a_d = a, b_d = b), by = c("distractor_label" = "item")) |>
  left_join(d_sub |> select(administration_id, prod), by = "administration_id") |>
  mutate(theta = cdi$prop_to_theta(prod),
         agg_k_t = -b_t, agg_k_d = -b_d,
         imp_k_t = p2pl(theta, a_t, b_t), imp_k_d = p2pl(theta, a_d, b_d))
```

```{r precise-signals}
items_long <- subset_dat$cdi_items_long |>
  mutate(item = str_to_lower(item), lab_subject_id = as.character(lab_subject_id)) |>
  rename(cdi_age = age)

admins <- base |>
  distinct(administration_id, subject_id, age) |>
  inner_join(subject_map, by = "subject_id") |>
  filter(lab_subject_id %in% items_long$lab_subject_id)

nearest_cdi <- admins |>
  inner_join(items_long |> distinct(lab_subject_id, cdi_age),
             by = "lab_subject_id", relationship = "many-to-many") |>
  mutate(age_diff = abs(age - cdi_age)) |>
  group_by(administration_id) |>
  slice_min(age_diff, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(administration_id, lab_subject_id, cdi_age)

produces_at_age <- items_long |>
  group_by(lab_subject_id, cdi_age, item) |>
  summarise(produces = as.integer(any(produces == 1)), .groups = "drop")

precise <- base |>
  inner_join(nearest_cdi, by = "administration_id") |>
  left_join(produces_at_age |> rename(prec_k_t = produces),
            by = c("lab_subject_id", "cdi_age", "target_label" = "item")) |>
  left_join(produces_at_age |> rename(prec_k_d = produces),
            by = c("lab_subject_id", "cdi_age", "distractor_label" = "item"))

cat("Precise subset: ", n_distinct(precise$lab_subject_id), " children, ",
    nrow(precise), " trials\n", sep = "")
```

## Four linking models, written out

For a trial by child *c* in dataset *s* with target/distractor knowledge
`k_t, k_d`, every model has the same random structure — a child intercept
`u_c` (absorbing general ability) and a dataset intercept `v_s` — and they differ
only in how knowledge enters the mean:

$$
\textbf{H0:}\quad \text{elogit} = \beta_0 + \beta_{age}\,\log\text{age} + u_c + v_s + \varepsilon
$$
$$
\textbf{H1:}\quad \ldots + \beta_t\,k_t \qquad\text{(target knowledge only)}
$$
$$
\textbf{H2:}\quad \ldots + \beta_t\,k_t + \beta_d\,k_d \qquad\text{(target and distractor, separately)}
$$
$$
\textbf{H3:}\quad \ldots + \beta_\Delta\,(k_t - k_d) \qquad\text{(symmetric 2AFC contrast)}
$$

Inferences we can draw:

- **H1 vs H0** — does knowing the target help at all? (`β_t > 0` expected.)
- **H2 vs H1** — does the *distractor* add anything beyond the target?
  (`β_d < 0` expected if a known competitor pulls looks away.)
- **H3** — forces target and distractor to matter *equally and oppositely*. If
  the true effect is asymmetric, H3 will fit *worse* than H2 even though it is
  the "tidy" 2AFC story. (This is exactly what we find — and notebook 6 explains
  why.)

Note these models include a child intercept but **not** a word intercept, so a
knowledge coefficient here captures the effect of word knowledge *including* the
fact that easier words are both better known and better recognized. Notebook 6
adds word intercepts to strip that out and isolate the purely item-specific part.

## How noisy is a trial? (the R² ceiling)

```{r noise-ceiling}
m_full <- lmer(elogit ~ log_age_centered + (1 | administration_id) + (1 | dataset_name),
               data = base, control = lmerControl(optimizer = "bobyqa"))
vc <- as.data.frame(VarCorr(m_full))
tibble(
  component = c("child differences", "dataset differences", "residual (trial noise)"),
  variance = c(vc$vcov[vc$grp == "administration_id"],
               vc$vcov[vc$grp == "dataset_name"],
               vc$vcov[vc$grp == "Residual"])
) |>
  mutate(share = variance / sum(variance)) |>
  kable(digits = c(0, 2, 3),
        caption = "Variance decomposition of trial elogit (age-only model)")
```

The residual (pure trial noise) is the large majority of the variance. That is
the ceiling: even a perfect knowledge model can only chip at the small
non-residual slice, so trial-level R² *must* be small. Read the numbers below as
relative, not absolute.

## Effect sizes: what the coefficients say

We fit H2 in each setting on the common precise subset (so the three settings
describe exactly the same trials) and report the knowledge coefficients in
logits and as accuracy effect sizes (evaluated at the sample's mean elogit).

```{r effect-sizes}
ref_elogit <- mean(precise$elogit, na.rm = TRUE)
pp <- function(beta) 100 * (plogis(ref_elogit + beta) - plogis(ref_elogit))

sub_frame <- precise |>
  filter(!is.na(prec_k_t), !is.na(prec_k_d), !is.na(agg_k_t), !is.na(agg_k_d),
         !is.na(imp_k_t), !is.na(imp_k_d))

fit_h2 <- function(d, kt, kd, label) {
  f <- as.formula(paste0("elogit ~ log_age_centered + ", kt, " + ", kd,
                         " + (1|administration_id) + (1|dataset_name)"))
  m <- lmer(f, data = d, control = lmerControl(optimizer = "bobyqa"))
  ci <- confint(m, parm = c(kt, kd), method = "Wald")
  tibble(setting = label,
         term = c("target (k_t)", "distractor (k_d)"),
         beta = fixef(m)[c(kt, kd)],
         ci_low = ci[, 1], ci_high = ci[, 2],
         acc_pp = pp(fixef(m)[c(kt, kd)]))
}

bind_rows(
  fit_h2(sub_frame, "agg_k_t",  "agg_k_d",  "aggregate"),
  fit_h2(sub_frame, "imp_k_t",  "imp_k_d",  "imputed"),
  fit_h2(sub_frame, "prec_k_t", "prec_k_d", "precise")
) |>
  kable(digits = 3,
        caption = "H2 knowledge coefficients (logit) and accuracy effect (pp) on the common subset")
```

For the binary **precise** signal the `acc_pp` for the target is the most direct
"how much does it matter" number: the accuracy gain for a word the child
*actually knows* vs one they don't, holding the child and dataset fixed. For the
**imputed/aggregate** signals the coefficient is per unit of the (continuous)
knowledge scale, so compare signs and significance rather than raw pp.

## Out-of-sample prediction (relative, against the noise ceiling)

Child-grouped 5-fold CV; predict held-out trials from fixed effects only
(`re.form = NA`), so we are scoring how well the *linking signal* generalizes to
new children. We report R² and, more usefully, the **improvement over the H0
baseline**.

```{r harness}
set.seed(42)
cv_r2 <- function(data, terms, n_folds = 5) {
  fixed <- paste(c("log_age_centered", terms), collapse = " + ")
  form <- as.formula(paste0("elogit ~ ", fixed,
                            " + (1|administration_id) + (1|dataset_name)"))
  subs <- unique(data$administration_id)
  fold <- setNames(sample(rep(1:n_folds, length.out = length(subs))), subs)
  data$.fold <- fold[as.character(data$administration_id)]
  preds <- rep(NA_real_, nrow(data))
  for (f in 1:n_folds) {
    tr <- data |> filter(.fold != f); te <- which(data$.fold == f)
    m <- tryCatch(lmer(form, data = tr, control = lmerControl(optimizer = "bobyqa")),
                  error = function(e) NULL)
    if (!is.null(m)) preds[te] <- predict(m, newdata = data[te, ], re.form = NA,
                                          allow.new.levels = TRUE)
  }
  ok <- !is.na(preds)
  if (sum(ok) < 10) return(NA_real_)
  cor(preds[ok], data$elogit[ok])^2
}

make_terms <- function(kt, kd) list(
  H0 = character(0), H1 = kt, H2 = c(kt, kd),
  H3 = paste0("I(", kt, " - ", kd, ")"))

run_setting <- function(data, kt, kd, label) {
  d <- data |> filter(!is.na(.data[[kt]]), !is.na(.data[[kd]]))
  terms <- make_terms(kt, kd)
  r2 <- map_dbl(terms, ~ cv_r2(d, .x))
  tibble(setting = label, hypothesis = names(terms), cv_r2 = r2,
         delta_vs_H0 = r2 - r2[["H0"]])
}
```

```{r compare-subset}
subset_results <- bind_rows(
  run_setting(sub_frame, "agg_k_t",  "agg_k_d",  "aggregate"),
  run_setting(sub_frame, "imp_k_t",  "imp_k_d",  "imputed"),
  run_setting(sub_frame, "prec_k_t", "prec_k_d", "precise")
)

subset_results |>
  select(setting, hypothesis, cv_r2) |>
  pivot_wider(names_from = hypothesis, values_from = cv_r2) |>
  kable(digits = 4, caption = paste0("Out-of-sample R² on the common subset (",
        n_distinct(sub_frame$lab_subject_id), " children, ", nrow(sub_frame), " trials)"))

subset_results |>
  filter(hypothesis != "H0") |>
  ggplot(aes(x = hypothesis, y = delta_vs_H0, fill = setting)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0) +
  ylab("R² improvement over age-only baseline") +
  ggtitle("Predictive gain from each linking model, by setting")
```

```{r compare-full}
full_results <- bind_rows(
  run_setting(base, "agg_k_t", "agg_k_d", "aggregate (full)"),
  run_setting(base |> filter(!is.na(theta)), "imp_k_t", "imp_k_d", "imputed (full)")
)
full_results |>
  select(setting, hypothesis, cv_r2) |>
  pivot_wider(names_from = hypothesis, values_from = cv_r2) |>
  kable(digits = 4, caption = "Out-of-sample R² on the full data")
```

## What to conclude

- **Word knowledge does predict accuracy** (H1 > H0 everywhere; the effect-size
  table quantifies it — a known target buys several points of accuracy).
- **The link is asymmetric.** H2 ≈ H1 (the distractor coefficient is ~0) and
  **H3 is no better than baseline**: forcing target and distractor to count
  equally (the tidy 2AFC contrast) throws the signal away. The decision is
  driven by *target* knowledge far more than by distractor competition.
- **Imputed ≈ precise here**, because this notebook's models contain no word
  intercept: the knowledge coefficient is allowed to ride on between-word
  difficulty, which imputation already encodes. The place where real item-level
  CDI must beat imputation is the *within-word* comparison — that is notebook 6.
- **Absolute R² is tiny and that is expected** (see the noise ceiling); the
  improvement-over-baseline and effect-size views are the honest summaries.
