---
title: "5. Linking hypotheses × data settings (unified evaluation)"
subtitle: "peekbank-items"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

This is the core notebook. We test a small set of **linking hypotheses** — ways
that a child's word knowledge could predict their trial-level LWL accuracy —
and evaluate each one across the available **data settings**, on a common metric
(cross-validated prediction of trial elogit, generalizing to held-out children).

**Data settings** (how word knowledge is represented):

- **aggregate** — word difficulty only, no child info: `k = -b` (IRT ease).
- **imputed** — `P(child knows word)` from the child's CDI-derived ability
  `theta` and the word's IRT params (large/noisy; all CDI children).
- **precise** — the child *actually* produced the word on the CDI (small/clean;
  the Stanford TotLot-cohort subset — `adams_marchman_2018`,
  `fernald_marchman_2012`, `fmw_2013` — with item-level CDI, matched via
  `lab_subject_id`).

**Linking hypotheses** (oriented so target knowledge `k_t` helps, distractor
knowledge `k_d` hurts):

- **H0** age only (baseline)
- **H1** target knowledge only: `+ k_t`
- **H2** additive: `+ k_t + k_d`
- **H3** 2AFC contrast: `+ (k_t - k_d)`  — the theory-derived predictor
  (for knowledge probabilities, `E[correct] = 0.5 + 0.5*(p_t - p_d)`)

The whole point of crossing hypotheses × settings is to ask: does the same
linking hypothesis win regardless of how knowledge is measured, and how much
does the precise (item-level) measurement buy us over the imputed one?

```{r setup}
library(here)
library(lme4)
source(here("helper", "common.R"))
knitr::opts_chunk$set(cache = FALSE)

d_trial <- readRDS(here("cached_intermediates", "1_d_trial.Rds")) |> ungroup()
d_sub   <- readRDS(here("cached_intermediates", "1_d_sub.Rds"))
cdi     <- readRDS(here("cached_intermediates", "3_cdi_difficulty.Rds"))
subset_dat <- readRDS(here("cached_intermediates", "0_cdi_items_subset.Rds"))

p2pl <- function(theta, a, b) plogis(a * (theta - b))
```

## Build the trial frame with knowledge signals

Start from trials with a usable elogit, join IRT difficulty for target and
distractor, and the child's CDI-derived ability.

```{r build-frame}
cdi_diff <- cdi$item_pars |> select(item, a, b)

base <- d_trial |>
  rename(accuracy = long_window_accuracy, elogit = long_window_elogit) |>
  filter(!is.na(elogit)) |>
  mutate(log_age = log(age), log_age_centered = log_age - mean(log_age)) |>
  # target IRT params
  left_join(cdi_diff |> rename(a_t = a, b_t = b), by = c("target_label" = "item")) |>
  # distractor IRT params
  left_join(cdi_diff |> rename(a_d = a, b_d = b), by = c("distractor_label" = "item")) |>
  # child ability (theta) from CDI prod proportion via the WS bridge
  left_join(d_sub |> select(administration_id, prod), by = "administration_id") |>
  mutate(theta = cdi$prop_to_theta(prod))

# aggregate signals (no child info): ease = -difficulty
base <- base |>
  mutate(
    agg_k_t = -b_t,
    agg_k_d = -b_d,
    # imputed signals: P(child knows word)
    imp_k_t = p2pl(theta, a_t, b_t),
    imp_k_d = p2pl(theta, a_d, b_d)
  )
```

### Precise subset signals (item-level CDI)

Bridge peekbank administrations to the Stanford item-level CDI via the
release-stable **`lab_subject_id`** (the link file's `peekbank_admin_id` is from a
2022 pull and does not match 2026.1). For each LWL administration we use the
child's CDI record **nearest in age**, then look up whether they produced each
target / distractor word.

```{r precise-signals}
subject_map <- readRDS(here("cached_intermediates", "0_subject_map.Rds"))
items_long <- subset_dat$cdi_items_long |>
  mutate(item = str_to_lower(item), lab_subject_id = as.character(lab_subject_id)) |>
  rename(cdi_age = age)

# per-trial child + age, mapped to lab_subject_id
admins <- base |>
  distinct(administration_id, subject_id, age) |>
  inner_join(subject_map, by = "subject_id") |>
  filter(lab_subject_id %in% items_long$lab_subject_id)

# choose, per administration, the CDI record nearest in age
cdi_ages <- items_long |> distinct(lab_subject_id, cdi_age)
nearest_cdi <- admins |>
  inner_join(cdi_ages, by = "lab_subject_id", relationship = "many-to-many") |>
  mutate(age_diff = abs(age - cdi_age)) |>
  group_by(administration_id) |>
  slice_min(age_diff, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(administration_id, lab_subject_id, cdi_age)

# produces at that chosen CDI age, per item. A child may have multiple CDI
# records at the same age (different forms / source files), so collapse to
# "produced on any record" to get one row per (child, age, item).
produces_at_age <- items_long |>
  group_by(lab_subject_id, cdi_age, item) |>
  summarise(produces = as.integer(any(produces == 1)), .groups = "drop")

precise <- base |>
  inner_join(nearest_cdi, by = "administration_id") |>
  left_join(produces_at_age |> rename(prec_k_t = produces),
            by = c("lab_subject_id", "cdi_age", "target_label" = "item")) |>
  left_join(produces_at_age |> rename(prec_k_d = produces),
            by = c("lab_subject_id", "cdi_age", "distractor_label" = "item"))

cat("Precise subset: ", n_distinct(precise$lab_subject_id), " children, ",
    nrow(precise), " trials; ",
    "target word matched a CDI item in ",
    round(mean(!is.na(precise$prec_k_t)) * 100), "% of trials\n", sep = "")
```

## Evaluation harness

For a given data frame and a set of fixed-effect terms, fit
`elogit ~ <age + terms> + (1|administration_id) + (1|dataset_name)` and evaluate
with **child-grouped 5-fold CV**, predicting with fixed effects only
(`re.form = NA`) so we measure how well the *linking signal* generalizes to new
children. Metric: out-of-sample R² (squared correlation of predicted vs observed
elogit), averaged over folds.

```{r harness}
set.seed(42)

cv_r2 <- function(data, terms, n_folds = 5) {
  data <- data |> filter(if_all(all_of(terms[terms %in% names(data)]), ~ !is.na(.x)))
  # build formula
  fixed <- paste(c("log_age_centered", terms), collapse = " + ")
  form <- as.formula(paste0("elogit ~ ", fixed,
                            " + (1|administration_id) + (1|dataset_name)"))
  # assign folds by child
  subs <- unique(data$administration_id)
  fold <- setNames(sample(rep(1:n_folds, length.out = length(subs))), subs)
  data$.fold <- fold[as.character(data$administration_id)]

  preds <- numeric(nrow(data)); preds[] <- NA
  for (f in 1:n_folds) {
    tr <- data |> filter(.fold != f)
    te <- which(data$.fold == f)
    m <- tryCatch(
      lmer(form, data = tr, control = lmerControl(optimizer = "bobyqa")),
      error = function(e) NULL)
    if (is.null(m)) next
    preds[te] <- predict(m, newdata = data[te, ], re.form = NA, allow.new.levels = TRUE)
  }
  ok <- !is.na(preds)
  if (sum(ok) < 10) return(NA_real_)
  cor(preds[ok], data$elogit[ok])^2
}

# hypotheses as term sets, parameterized by a setting's knowledge column names
make_terms <- function(kt, kd) list(
  H0 = character(0),
  H1 = kt,
  H2 = c(kt, kd),
  H3 = paste0("I(", kt, " - ", kd, ")")
)

run_setting <- function(data, kt, kd, label) {
  # need both knowledge columns present
  d <- data |> filter(!is.na(.data[[kt]]), !is.na(.data[[kd]]))
  terms <- make_terms(kt, kd)
  tibble(
    setting = label,
    hypothesis = names(terms),
    n_trials = nrow(d),
    cv_r2 = map_dbl(terms, ~ cv_r2(d, .x))
  )
}
```

## Apples-to-apples: all settings on the precise subset

The precise signals exist only for the Stanford TotLot-cohort subset, so for a
fair three-way comparison we run **all three settings on exactly these trials**.

```{r compare-subset}
sub_frame <- precise |>
  filter(!is.na(prec_k_t), !is.na(prec_k_d),
         !is.na(agg_k_t), !is.na(agg_k_d),
         !is.na(imp_k_t), !is.na(imp_k_d))

subset_results <- bind_rows(
  run_setting(sub_frame, "agg_k_t",  "agg_k_d",  "aggregate"),
  run_setting(sub_frame, "imp_k_t",  "imp_k_d",  "imputed"),
  run_setting(sub_frame, "prec_k_t", "prec_k_d", "precise")
)

subset_results |>
  select(setting, hypothesis, cv_r2) |>
  pivot_wider(names_from = hypothesis, values_from = cv_r2) |>
  kable(digits = 4, caption = paste0("CV R2 on the precise subset (",
        n_distinct(sub_frame$lab_subject_id), " children, ", nrow(sub_frame), " trials)"))
```

```{r compare-subset-delta}
subset_results |>
  group_by(setting) |>
  mutate(delta_vs_H0 = cv_r2 - cv_r2[hypothesis == "H0"]) |>
  filter(hypothesis != "H0") |>
  ggplot(aes(x = hypothesis, y = delta_vs_H0, fill = setting)) +
  geom_col(position = "dodge") +
  ylab("CV R2 improvement over age-only baseline") +
  ggtitle("Linking hypotheses x data settings (precise subset)")
```

## Power run: aggregate and imputed on the full data

```{r compare-full}
full_results <- bind_rows(
  run_setting(base, "agg_k_t", "agg_k_d", "aggregate (full)"),
  run_setting(base |> filter(!is.na(theta)), "imp_k_t", "imp_k_d", "imputed (full)")
)

full_results |>
  select(setting, hypothesis, n_trials, cv_r2) |>
  pivot_wider(names_from = hypothesis, values_from = cv_r2) |>
  kable(digits = 4, caption = "CV R2 on full data (aggregate and imputed settings)")
```

## Read

The subset table is the key comparison: reading **down a column** shows whether a
given linking hypothesis improves as knowledge measurement sharpens
(aggregate → imputed → precise); reading **across a row** shows which hypothesis
form (target-only vs additive vs 2AFC contrast) wins. The full-data table checks
that the winning pattern holds with real power on the imputed pathway that scales
to the whole database.
