---
title: "6. Item-specific knowledge: within-child, within-word linkage"
subtitle: "peekbank-items"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

Notebook 5 showed that most of the trial-level predictive signal is a **child
general-ability main effect**: the 2AFC contrast (H3) added nothing, and the
precise (item-level) pathway did not beat the imputed one. That is the expected
result *if* the only thing being measured is "smarter kids look more at targets."

Here we ask the sharper question the prior literature never isolated:

> Controlling for a child's general ability **and** for a word's general
> difficulty/salience, does knowing *this specific word* predict accuracy on
> *this specific trial*?

The design that isolates this is a crossed random-effects model. With a child
random intercept (absorbing general ability) and target/distractor word random
intercepts (absorbing general word difficulty and salience), a `produces`
**fixed effect** is identified only from *within-child, within-word* variation —
the same child knowing word A but not word B, and the same word known by child X
but not child Y. That coefficient is the item-specific knowledge effect.

A key conceptual point falls out immediately: the **imputed** knowledge signal
`P(knows) = 2PL(theta_child, a_word, b_word)` is a deterministic function of
child ability and word difficulty, so once child and word effects are in the
model it carries **no** additional information. Only **real item-level CDI**
(the precise subset) has residual item-specific variance. This notebook is
therefore necessarily about the precise subset.

```{r setup}
library(here)
library(lme4)
source(here("helper", "common.R"))
knitr::opts_chunk$set(cache = FALSE)

d_trial <- readRDS(here("cached_intermediates", "1_d_trial.Rds")) |> ungroup()
d_sub   <- readRDS(here("cached_intermediates", "1_d_sub.Rds"))
cdi     <- readRDS(here("cached_intermediates", "3_cdi_difficulty.Rds"))
subset_dat  <- readRDS(here("cached_intermediates", "0_cdi_items_subset.Rds"))
subject_map <- readRDS(here("cached_intermediates", "0_subject_map.Rds"))

p2pl <- function(theta, a, b) plogis(a * (theta - b))
```

## Build the precise frame

Same construction as notebook 5: trial elogit, IRT params for target/distractor,
child ability `theta`, and the item-level `produces` (nearest-age CDI, bridged on
`lab_subject_id`).

```{r build}
cdi_diff <- cdi$item_pars |> select(item, a, b)

base <- d_trial |>
  rename(accuracy = long_window_accuracy, elogit = long_window_elogit) |>
  filter(!is.na(elogit)) |>
  mutate(log_age = log(age), log_age_centered = log_age - mean(log_age)) |>
  left_join(cdi_diff |> rename(a_t = a, b_t = b), by = c("target_label" = "item")) |>
  left_join(cdi_diff |> rename(a_d = a, b_d = b), by = c("distractor_label" = "item")) |>
  left_join(d_sub |> select(administration_id, prod), by = "administration_id") |>
  mutate(theta = cdi$prop_to_theta(prod))

items_long <- subset_dat$cdi_items_long |>
  mutate(item = str_to_lower(item), lab_subject_id = as.character(lab_subject_id)) |>
  rename(cdi_age = age)

admins <- base |>
  distinct(administration_id, subject_id, age) |>
  inner_join(subject_map, by = "subject_id") |>
  filter(lab_subject_id %in% items_long$lab_subject_id)

nearest_cdi <- admins |>
  inner_join(items_long |> distinct(lab_subject_id, cdi_age),
             by = "lab_subject_id", relationship = "many-to-many") |>
  mutate(age_diff = abs(age - cdi_age)) |>
  group_by(administration_id) |>
  slice_min(age_diff, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(administration_id, lab_subject_id, cdi_age)

produces_at_age <- items_long |>
  group_by(lab_subject_id, cdi_age, item) |>
  summarise(produces = as.integer(any(produces == 1)), .groups = "drop")

precise <- base |>
  inner_join(nearest_cdi, by = "administration_id") |>
  left_join(produces_at_age |> rename(prec_k_t = produces),
            by = c("lab_subject_id", "cdi_age", "target_label" = "item")) |>
  left_join(produces_at_age |> rename(prec_k_d = produces),
            by = c("lab_subject_id", "cdi_age", "distractor_label" = "item")) |>
  filter(!is.na(prec_k_t), !is.na(prec_k_d))

cat(nrow(precise), "trials,", n_distinct(precise$lab_subject_id), "children,",
    n_distinct(precise$target_label), "target words\n")
```

## A. There IS item-specific variance to find

Decompose the real item-level knowledge into the part predictable from child
ability × word difficulty (what imputation captures) and the **residual** (what
imputation cannot capture). If the residual variance is ~0, item-level CDI buys
nothing; if it is substantial, there is item-specific signal to test.

```{r residual-variance}
res <- precise |>
  filter(!is.na(theta), !is.na(a_t), !is.na(b_t)) |>
  mutate(expected_t = p2pl(theta, a_t, b_t),
         residual_t = prec_k_t - expected_t)

tibble(
  quantity = c("Var(actual produces_target)",
               "Var(expected from ability x difficulty)",
               "Var(residual = item-specific)",
               "share item-specific"),
  value = c(var(res$prec_k_t),
            var(res$expected_t),
            var(res$residual_t),
            var(res$residual_t) / var(res$prec_k_t))
) |> kable(digits = 3)
```

A large residual share means children's word-by-word knowledge departs
substantially from what their overall vocabulary size predicts — exactly the
signal a child×item linkage needs.

## B. The within-child, within-word test

Nested crossed random-effects models. Child (`administration_id`) and word
(`target_label`, `distractor_label`) random intercepts absorb general ability and
general word difficulty/salience; the `produces` **fixed effects** are then the
item-specific knowledge effects.

```{r models}
ctrl <- lmerControl(optimizer = "bobyqa")
re <- "(1 | administration_id) + (1 | dataset_name) + (1 | target_label) + (1 | distractor_label)"

m0 <- lmer(as.formula(paste("elogit ~ log_age_centered +", re)),
           data = precise, control = ctrl)
m1 <- lmer(as.formula(paste("elogit ~ log_age_centered + prec_k_t +", re)),
           data = precise, control = ctrl)
m2 <- lmer(as.formula(paste("elogit ~ log_age_centered + prec_k_t + prec_k_d +", re)),
           data = precise, control = ctrl)

anova(m0, m1, m2) |> kable(digits = c(0, 1, 1, 1, 2, 0, 4),
  caption = "Does item-specific target/distractor knowledge improve fit?")
```

Fixed-effect estimates from the full model (these are the within-child,
within-word item-specific effects):

```{r effects}
ci <- confint(m2, parm = c("prec_k_t", "prec_k_d"), method = "Wald")
fe <- fixef(m2)[c("prec_k_t", "prec_k_d")]
tibble(
  term = c("knows target (prec_k_t)", "knows distractor (prec_k_d)"),
  estimate = fe,
  ci_low = ci[, 1],
  ci_high = ci[, 2]
) |> kable(digits = 3,
  caption = "Item-specific knowledge effects on trial elogit (target should be +, distractor -)")
```

## C. Descriptive: within-child known vs unknown

The cleanest picture: within each child, compare mean accuracy on trials whose
target word they **did** vs **did not** produce on the CDI. Differencing within
child removes general ability entirely.

```{r within-child}
wc <- precise |>
  group_by(lab_subject_id, prec_k_t) |>
  summarise(acc = mean(accuracy), n = n(), .groups = "drop") |>
  mutate(knows_target = ifelse(prec_k_t == 1, "knows", "does not know")) |>
  filter(n >= 3)  # need a few trials per cell

wc_paired <- wc |>
  select(lab_subject_id, knows_target, acc) |>
  pivot_wider(names_from = knows_target, values_from = acc) |>
  filter(!is.na(knows), !is.na(`does not know`)) |>
  mutate(diff = knows - `does not know`)

ggplot(wc_paired, aes(x = `does not know`, y = knows)) +
  geom_abline(lty = 2) +
  geom_point(alpha = .5) +
  coord_equal() +
  xlab("Accuracy on UNKNOWN target words") +
  ylab("Accuracy on KNOWN target words") +
  labs(subtitle = sprintf(
    "%d children with both cells; mean within-child advantage = %.3f (paired t p = %.4f)",
    nrow(wc_paired), mean(wc_paired$diff),
    t.test(wc_paired$knows, wc_paired$`does not know`, paired = TRUE)$p.value))
```

## Read

The notebook 5 result said the *aggregate* predictive signal is dominated by
general ability. This notebook asks the residual question that actually defines
an item-level linkage. Three things to read together:

- **A** quantifies whether real item-level CDI even contains item-specific
  variance beyond what vocabulary size implies (it must, for any of this to
  work — and it bounds the ceiling for the imputed pathway, which has *none*).
- **B** is the inferential test: a positive `prec_k_t` (and negative `prec_k_d`)
  with the child and word effects partialled out is direct evidence that
  knowing the specific words drives the specific trial — the 2AFC decision the
  background analyses argued was being missed.
- **C** is the assumption-light version of the same test.

If B/C are positive, the linkage *does* exist at the item level once measured
properly — it was just swamped by ability and salience in every aggregate
analysis. If they are null even here, the pessimistic prior conclusion holds
even with gold-standard item-level CDI.
