---
title: "6. Item-specific knowledge: within-child, within-word, over time"
subtitle: "peekbank-items"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

Notebook 5 established that word knowledge predicts looking accuracy, but the
predictive signal there can ride on two confounds: some **children** are just
better, and some **words** are just easier (more salient, earlier learned). The
question that actually defines an item-level linkage strips both away:

> Holding the child and the word fixed, does knowing *this* word predict accuracy
> on *this* word?

We answer it two ways, the second of which uses the subset's most valuable
feature — it is **longitudinal**.

1. **Cross-sectional, within child & within word** — crossed random effects pull
   out general child ability and general word difficulty; the knowledge
   coefficient is what's left.
2. **Longitudinal, within child×word over time** — for the 600+ child–word pairs
   where the child *learns the word* across visits, ask whether accuracy on that
   word rises when they acquire it, using each pair as its own control and
   separating word-learning from general maturation.

Because imputed knowledge `P(knows) = 2PL(θ_child, a_word, b_word)` is a
deterministic function of child ability and word difficulty, it carries **zero**
information once child and word are held fixed. So this test is only possible
with the **real item-level CDI** in the TotLot subset.

```{r setup}
library(here)
library(lme4)
source(here("helper", "common.R"))
knitr::opts_chunk$set(cache = FALSE)

d_trial <- readRDS(here("cached_intermediates", "1_d_trial.Rds")) |> ungroup()
d_sub   <- readRDS(here("cached_intermediates", "1_d_sub.Rds"))
cdi     <- readRDS(here("cached_intermediates", "3_cdi_difficulty.Rds"))
subset_dat  <- readRDS(here("cached_intermediates", "0_cdi_items_subset.Rds"))
subject_map <- readRDS(here("cached_intermediates", "0_subject_map.Rds"))

p2pl <- function(theta, a, b) plogis(a * (theta - b))
# elogit -> accuracy, and an effect in pp evaluated near the sample mean
acc <- function(el) plogis(el)
```

## Build the precise frame

Same construction as notebook 5 (item-level CDI bridged on `lab_subject_id`,
nearest CDI in age), plus a child×word pair id `cw` for the longitudinal models.

```{r build}
cdi_diff <- cdi$item_pars |> select(item, a, b)

base <- d_trial |>
  rename(accuracy = long_window_accuracy, elogit = long_window_elogit) |>
  filter(!is.na(elogit)) |>
  mutate(log_age = log(age), log_age_centered = log_age - mean(log_age)) |>
  left_join(cdi_diff |> rename(a_t = a, b_t = b), by = c("target_label" = "item")) |>
  left_join(cdi_diff |> rename(a_d = a, b_d = b), by = c("distractor_label" = "item")) |>
  left_join(d_sub |> select(administration_id, prod), by = "administration_id") |>
  mutate(theta = cdi$prop_to_theta(prod))

items_long <- subset_dat$cdi_items_long |>
  mutate(item = str_to_lower(item), lab_subject_id = as.character(lab_subject_id)) |>
  rename(cdi_age = age)

admins <- base |>
  distinct(administration_id, subject_id, age) |>
  inner_join(subject_map, by = "subject_id") |>
  filter(lab_subject_id %in% items_long$lab_subject_id)

nearest_cdi <- admins |>
  inner_join(items_long |> distinct(lab_subject_id, cdi_age),
             by = "lab_subject_id", relationship = "many-to-many") |>
  mutate(age_diff = abs(age - cdi_age)) |>
  group_by(administration_id) |>
  slice_min(age_diff, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(administration_id, lab_subject_id, cdi_age)

produces_at_age <- items_long |>
  group_by(lab_subject_id, cdi_age, item) |>
  summarise(produces = as.integer(any(produces == 1)), .groups = "drop")

precise <- base |>
  inner_join(nearest_cdi, by = "administration_id") |>
  left_join(produces_at_age |> rename(prec_k_t = produces),
            by = c("lab_subject_id", "cdi_age", "target_label" = "item")) |>
  left_join(produces_at_age |> rename(prec_k_d = produces),
            by = c("lab_subject_id", "cdi_age", "distractor_label" = "item")) |>
  filter(!is.na(prec_k_t), !is.na(prec_k_d)) |>
  mutate(cw = paste(lab_subject_id, target_label, sep = "_"))

ref_elogit <- mean(precise$elogit)
pp <- function(beta) 100 * (acc(ref_elogit + beta) - acc(ref_elogit))  # effect in accuracy pp

n_visits <- precise |> distinct(lab_subject_id, administration_id) |> count(lab_subject_id)
cat(sprintf("%d trials, %d children, %d target words; median %g visits/child (max %d) -- longitudinal\n",
            nrow(precise), n_distinct(precise$lab_subject_id),
            n_distinct(precise$target_label), median(n_visits$n), max(n_visits$n)))
```

## A. Is there item-specific variance to find?

Split each child's real per-word knowledge into the part predictable from their
overall vocabulary × the word's difficulty (what imputation captures) and the
**residual** (genuinely item-specific). If the residual is ~0, item-level CDI
adds nothing over imputation.

```{r residual-variance}
res <- precise |>
  filter(!is.na(theta), !is.na(a_t), !is.na(b_t)) |>
  mutate(expected_t = p2pl(theta, a_t, b_t), residual_t = prec_k_t - expected_t)

tibble(
  quantity = c("Var(actual knows-target)",
               "Var(expected from ability × difficulty)",
               "Var(residual = item-specific)",
               "item-specific share"),
  value = c(var(res$prec_k_t), var(res$expected_t),
            var(res$residual_t), var(res$residual_t) / var(res$prec_k_t))
) |> kable(digits = 3)
```

A large item-specific share means children's word-by-word knowledge departs a lot
from what vocabulary size alone predicts — there is real item-level signal, and
it sets the ceiling that imputation (item-specific share = 0) can never reach.

## B. Cross-sectional test, spelled out

Three nested models. **All** share the same random effects, which is the whole
point — each one absorbs a confound so the fixed effect of knowledge is read off
the variation that remains:

- `(1 | administration_id)` — the **child's** general level (absorbs ability)
- `(1 | dataset_name)` — study differences
- `(1 | target_label)` — the **target word's** general difficulty/salience
- `(1 | distractor_label)` — the **distractor word's** general grabbiness

$$
\textbf{m0:}\quad \text{elogit} = \beta_0 + \beta_{age}\log\text{age} + u_{child} + u_{tword} + u_{dword} + v_{dataset} + \varepsilon
$$
$$
\textbf{m1:}\quad \ldots + \beta_t\,(\text{knows target}_{child,word})
$$
$$
\textbf{m2:}\quad \ldots + \beta_t\,(\text{knows target}) + \beta_d\,(\text{knows distractor})
$$

Because the child and word means are already in the model, **`β_t` is identified
only from within-child, within-word variation**: the same child knowing word A
but not word B, and the same word known by some children but not others. That is
precisely the item-specific effect. Inferences:

- **m1 vs m0** — does item-specific *target* knowledge matter at all?
- **m2 vs m1** — does item-specific *distractor* knowledge add anything?

```{r models}
ctrl <- lmerControl(optimizer = "bobyqa")
re <- "(1|administration_id) + (1|dataset_name) + (1|target_label) + (1|distractor_label)"
m0 <- lmer(as.formula(paste("elogit ~ log_age_centered +", re)), precise, control = ctrl)
m1 <- lmer(as.formula(paste("elogit ~ log_age_centered + prec_k_t +", re)), precise, control = ctrl)
m2 <- lmer(as.formula(paste("elogit ~ log_age_centered + prec_k_t + prec_k_d +", re)), precise, control = ctrl)

anova(m0, m1, m2) |>
  kable(digits = c(0,1,1,1,2,0,4), caption = "Nested model comparison (m0 → m1 → m2)")
```

```{r effects}
ci <- confint(m2, parm = c("prec_k_t", "prec_k_d"), method = "Wald")
tibble(
  term = c("knows target", "knows distractor"),
  beta_logit = fixef(m2)[c("prec_k_t", "prec_k_d")],
  ci_low = ci[, 1], ci_high = ci[, 2],
  accuracy_pp = pp(fixef(m2)[c("prec_k_t", "prec_k_d")])
) |>
  kable(digits = 3,
        caption = "Item-specific effects: logit, 95% CI, and accuracy (pp). Target should be +, distractor ~0/−.")
```

The **accuracy_pp** column is the headline effect size: how many percentage
points of looking accuracy a child gains on a word they specifically know, with
their overall ability and the word's difficulty held fixed.

## C. Longitudinal test: does learning a word raise accuracy on it?

The subset is longitudinal (median several visits per child), so for many
child×word pairs we see the child **before and after** they acquire the word on
the CDI. Adding a child×word random intercept `(1 | cw)` means `β_t` is now
identified *within pair, across time*: the pair is its own control, so any stable
reason a given child is good at a given word is differenced out. Keeping `log age`
in the model separates **word learning** from **general maturation** (kids get
better at everything with age).

$$
\text{elogit} = \beta_0 + \beta_{age}\log\text{age} + \beta_t\,(\text{knows target}_{child,word,\,time}) + u_{tword} + u_{cw} + \varepsilon
$$

```{r longitudinal-counts}
pair_summary <- precise |>
  group_by(cw) |>
  summarise(n_ages = n_distinct(age), changes = n_distinct(prec_k_t), .groups = "drop")
cat(sprintf("child×word pairs: %d total; %d seen at ≥2 ages; %d show the child learning the word (0→1)\n",
            nrow(pair_summary), sum(pair_summary$n_ages >= 2),
            sum(pair_summary$n_ages >= 2 & pair_summary$changes == 2)))
```

```{r longitudinal-model}
mL0 <- lmer(elogit ~ log_age_centered + (1|target_label) + (1|cw), precise, control = ctrl)
mL1 <- lmer(elogit ~ log_age_centered + prec_k_t + (1|target_label) + (1|cw), precise, control = ctrl)

ciL <- confint(mL1, parm = "prec_k_t", method = "Wald")
cat(sprintf("Within child×word effect of learning the target: %.3f logit [%.3f, %.3f]  = %+.1f pp accuracy\n",
            fixef(mL1)["prec_k_t"], ciL[1], ciL[2], pp(fixef(mL1)["prec_k_t"])))
anova(mL0, mL1) |> kable(digits = c(0,1,1,1,2,0,4),
  caption = "Likelihood-ratio test for the within-pair, over-time learning effect")
```

This estimate is the strongest evidence for the linkage: it comes only from
moments where a child's knowledge of a specific word changed, with the word's
identity, the child's identity, and general age effects all controlled.

### Before vs after acquisition

For the pairs where the child learns the word, mean accuracy on that word while
**not yet** producing it vs **after** they produce it (each point is one
child×word pair).

```{r before-after}
learners <- pair_summary |> filter(n_ages >= 2, changes == 2) |> pull(cw)
ba <- precise |>
  filter(cw %in% learners) |>
  group_by(cw, prec_k_t) |>
  summarise(acc = mean(accuracy), .groups = "drop") |>
  mutate(state = ifelse(prec_k_t == 1, "after", "before")) |>
  select(cw, state, acc) |>
  pivot_wider(names_from = state, values_from = acc) |>
  filter(!is.na(before), !is.na(after))

tt <- t.test(ba$after, ba$before, paired = TRUE)
ggplot(ba, aes(x = before, y = after)) +
  geom_abline(lty = 2) +
  geom_point(alpha = .4) +
  coord_equal() +
  xlab("Accuracy BEFORE acquiring the word") +
  ylab("Accuracy AFTER acquiring the word") +
  labs(subtitle = sprintf("%d learner pairs; mean gain %+.1f pp (paired t p = %.4g)",
                          nrow(ba), 100 * mean(ba$after - ba$before), tt$p.value))
```

## What to conclude

- **The item-level linkage is real.** Holding child ability and word difficulty
  fixed, knowing the specific target word raises looking accuracy by several
  percentage points (Section B), and the effect reproduces **within child×word
  over time**: when a child learns a word, accuracy on that word goes up beyond
  what maturation explains (Section C).
- **It is asymmetric** — target knowledge drives it, distractor knowledge does
  not. This is the concrete reason the symmetric 2AFC contrast (H3 in notebook 5)
  failed, and why aggregate analyses, which average over both, saw nothing.
- **Real item-level CDI is necessary.** Two-thirds of word-by-word knowledge is
  item-specific residual (Section A) that imputation cannot represent, so this
  result is invisible to the full-sample imputed pathway by construction.

The prior "no item-level correlation" conclusion was a measurement-and-modeling
artifact: the signal was swamped by child ability and word salience, and erased
by averaging. With longitudinal item-level CDI and a within-pair model, it is
clearly present.
